National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs

### 2020-11-16 Manuscript.Rmd
Adding figures, updating manuscript (clarified introduction)

---
title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
shorttitle        : "Beliefs and Wellbeing"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    address       : "Postal address"
    email         : "my@email.com"
    role:         # First author
      - Conceptualization
      - Literature Review
      - Writing - First draft
      - Writing - Review & editing
  - name          : "Ev l Worthing"
    affiliation   : "2"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Don E Davis"
    affiliation   : "3"
    address       : "Postal address"
    email         : "my@email.com"
    role:         
      - Writing review & editing
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    address       : "Postal address"
    email         : "my@email.com"
    role:        
      - Writing review & editing
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    role:         
      - Data analysis
      - Writing second and subsequent drafts
      - Writing review & editing

authornote: |
  Add complete departmental affiliations for each author here. Each new line herein must be indented, like this line.
  Enter author note here.
abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and the reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel response data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018). Contrary to previous studies, we find that belief in a spirit or life force predicts initially lower personal wellbeing and life-satisfaction. However, in support of theories that sprituality improves psychological health, beliefs in a spirit or life force predict increasing life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences. Intriguingly these findings also hold among those who believe in a God but disbelieve in a spirit or life force. The causal mechanisms by which beliefs in a spirit or life force increases well-being remain an important horizon for future investigations.

  <!-- https://tinyurl.com/ybremelq -->
keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : no
bibliography      : ["b-references.bib"]
floatsintext      : no
figurelist        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no
documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
---

@@

```{r load_libraries, echo=FALSE, include=FALSE}
# load libraries
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("sjPlot")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("gghighlight")
library("sjPlot")
library("see")
library("viridis")
library("ggpubr")
library("wesanderson")
library("ggsci")
library("here")
library("citr")
#library("furniture")
library("knitr")
library("texreg")
```
```{r load_data, include=FALSE}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```
```{r data processing, cache = TRUE, dependson = "ldf.5"}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

@@

The relationship between spirituality and psychological well-being is a matter of enduring human fascination [@Ellsworth2010-yu]. Previous attempts to quantify the relationship between these two domains find a positiven association between spirituality and well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics have raised three challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that well-being predicts well-being –. what Koenig describes as a  "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies investigating spirituality and emotional well-being operationalise spirituality scales using constructs for positive emotional well-being [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

How, then, should researchers operationalise spirituality. We can look to folk traditions for guidance. The term "spirituality" was first used to describe the religious practices of ascetics and monks. more recently, however, the term has taken on a family of broader meaning [@Swinton2001-vr]. Religion is typically operationalized to mean beliefs and practices respecting a God or gods (JB). A recurring element in the use of ‘spirituality,’ as it is distinguished from religion, is belief in a spirit or life force without commitment to a belief in a God or gods [@Hill2000-cl; @Davis2015-tx]. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force. We operationalize spiritual belief as either belief in a spirit or life force without any corresponding belief in a God. However, the term "spirituality" overlaps with the uses of "religion" in the sense that God or gods are instances of beliefs in spirits – that is one may be both religious and spiritual. Here we operationalize religious belief as either belief in a God or belief in both a God and a spirit or life force.  isNotably, previous spirituality mental health research has been disproportionately conducted on North American samples. In the United States, especially, samples include a majority of people who believe in both a God as well as and life forces other than God [@Smith2003-re; @Koenig2010-gk]. Although such studies drawing from such samples mightmay clarify the effects of religious spirituality on psychological subjective well-being, the question of how non-religious spirituality affects psychological well-being that is – or spiritual beliefs without any belief in a God or gods – affects subjective well-being remain unclear.  (For discussion, see: Therefore, to better understand possible distinctions between religious spirituality and non-religious spirituality for subjective well-being outcomes, samples collected from cultures outside the North Atlantic that contain a more equal ratio of the two are needed [@Garssen2016-km]).

Third, most research investigating the relationship between spirituality and subjective well-being has relied on cross-sectional samples. These studies include high-quality reviews and meta-analyses [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. However, to assess whetherto whether spiritual beliefs lead to a growth, stability, or decline in psychological well-being relative to people who lack such beliefs , including people who believe in God but not a spirit or lifeforce, requires repeatedly measuring the same individuals over time. clarify causal effects requires measuring the same individuals over time and assessing how life turns out for demographically similar people who differ in their spiritual and religious beliefs over their lifespans  and orientations [@Garssen2016-km].

Our study addresses these three limitations in previous research.improves on past methods the following ways: First, we avoid a tautological elision of "spirituality" and "well-being"y by adopting a non-affective measure of spirituality through use of general, non-affective measures: "Do you believe in some form of a spirit or life force?" Our measures of psychological well being are [BEN SAY HERE] and Second we adopt a four-level categorical indicator that disentangles spiritual and religious beliefs: (1) Skeptics: those who neither believe in a God nor a life force; (2) SpiritExcludeGod: those who no not believe in a God  but believe in a spirit or life force; (3) GodAndSpirit: those who believe in a God and spirit; (4) GodExcludesSpirit: those who believe in both a God and a spirit or life force "Do you believe in a god?". Second, we address the prior dependence on cross-sectional samples taken from the United States by analyzing data taken from By sampling from respondents in New Zealand, a country with [GIVE NON_RELIGIOUS SPIRITUAL STATS HERE).  Third, we are able to assess whether spirit beliefs are associate with a growth in psychological wellbeing address causation by weusing we use longitudinal responses from athis nationally diverse probability sample year over an nine-yearseight-year period with nine measurement points  to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-beingwellbeing (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

<!-- We rigorously assess the relationship between our non-affective measures and two constructs of subjective well-being (personal well-being and life satisfaction), adjusting for known demographic and personal influences, including age, sex, deprivation, and political orientation (Okulicz-Kozaryn, Holmes, and Avery 2014; Cuesta and Budría 2012; Røysamb et al. 2002; Hansen and Slagsvold 2012; Chen and van Ours, n.d.; Vaughan et al. 1985; Easterlin, Angelescu, and Zweig 2011; Stam et al. 2016; Yakovlev and Leguizamon 2012). Collectively, then, our approach avoids the limitations of many previous studies and helps to clarify existing literature concerning how best to study the relationship between spirituality and subjective well-being. -->

## Ethics Statement

@@

**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index (Cummins et al. 2017). Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

Our findings support critiques of the reliability of past methods. The majority of studies have found positive relationships between religion and spirituality and subjective well-being. These previous studies primarily use samples taken from within the North Atlantic region and are cross-sectional in design (Ano and Vasconcelles 2005; Hackney and Sanders 2003; Sawatzky, Ratner, and Chiu 2005; Smith, McCullough, and Poll 2003; Visser, Garssen, and Vingerhoets 2010; Yonker, Schnabelrauch, and Dehaan 2012). However, our results find a negative association between both belief in a spirit or life force that excludes a god (BXG) and belief that includes a god (BIG). These longitudinal results persist independent of time. Notably, our findings do partially support other previous work that shows that spirituality outside of a religious framework is associated with worse emotional well-being, though we also find a negative association within a religious framework (King et al. 2013). Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.

>>>>

---
title             : "National Longitudinal Study Reveals Growth in Wellbeing Among Those with Spiritual Beliefs"
shorttitle        : "Beliefs and Wellbeing"
author: 
  - name          : "Ben Highland"
    affiliation   : "1"
    corresponding : yes    # Define only one corresponding author
    email         : "benhighland6420@gmail.com"
  - name          : "Everett L. Worthington"
    affiliation   : "2"
    institution   : "Department of Psychology Virginia Commonwealth University"
  - name          : "Don E Davis"
    affiliation   : "3"
    institution   : "Department of Psychology Georgia State University"
  - name          : "Chris G. Sibley"
    affiliation   : "4"
    institution   : "School of Psychology University of Auckland"
  - name          : "Joseph Bulbulia"
    affiliation   : "5"
    institution   : "School of Psychology Victoria University"

authornote: |
  1 Wake Forest Medical School 2 Department of Psychology Virginia Commonwealth University 3 Department of Psychology Georgia State University 4 School of Psychology University of Auckland 5 School of Psychology Victoria University

abstract: |
 Previous research finds an association between spirituality and subjective well-being. However, the widespread use of tautological scales and reliance on cross-sectional data cast doubts. Here, we leverage nine years of panel data from a nationally diverse longitudinal study to systematically test whether having spiritual beliefs leads to growth in personal well-being and life satisfaction ($N$ =  20979, New Zealand, 2010-2018). We find that belief in a spirit or life force predicts lower personal wellbeing and life-satisfaction. However, in support of previous theories, beliefs in a spirit or life force predict increasing personal well-being and life satisfaction over time relative to disbelief. These finding holds after adjusting for known demographic influences, and intriguingly, among those who believe in a God but disbelieve in a spirit or life force. The recent growth in spiritual beliefs and decline in traditional religion across many industrial societies motivates futher causal nvestigations of the social and psychological mechanisms through which spiritual beliefs lead to growth in subjective wellbeing. 
  <!-- https://tinyurl.com/ybremelq -->

keywords          : "Belief, God, Health, Longitudinal, Panel, Religion, Spirit, Spirituality, Well-being"
wordcount         : "4778"
bibliography      : ["b-references.bib"]

floatsintext      : no
figurelist        : no
figsintext        : no
tablelist         : no
footnotelist      : no
linenumbers       : no
mask              : no
draft             : no

documentclass     : "apa6"
classoption       : "man"
output            : papaja::apa6_pdf
extra_dependencies: ["rotating", "booktabs", "graphicx","longtable"]
header-includes:
   - \usepackage{tcolorbox}
   - \usepackage{paralist}
   - \usepackage{multicol}
   - \usepackage{graphics}
   - \usepackage{rotating}
   - \usepackage{longtable}
---

@@

```{r load_libraries, echo=FALSE, include=FALSE}
# load libraries
library("styler")
library("tidyverse")
library("lme4")
library("ggplot2")
library("tidyr")
library("sjPlot")
library("ggeffects")
library("Amelia")
library("table1")
library("patchwork")
library("parameters")
library("prettycode")
library("gghighlight")
library("sjPlot")
library("see")
library("viridis")
library("ggpubr")
library("wesanderson")
library("ggsci")
library("here")
library("citr")
#library("furniture")
library("knitr")
library("texreg")
```
```{r load_data, include=FALSE}
ldf.5<-readRDS("~/The\ Virtues\ Project\ Dropbox/Joseph\ Bulbulia/00Bulbulia\ Pubs/2020/ldf.5") 
```
```{r data processing, cache = TRUE, dependson = "ldf.5", include=FALSE}
# filter only those who are measured 
dfben  <-ldf.5 %>%
  dplyr:: filter(Wave !=2009)%>% # exclude wave 1 (no spirit measures that wave)
  dplyr::filter(YearMeasured==1)%>%
  dplyr::group_by(Id) %>% filter(n() > 2)%>% # select those who have responded to at least 3 waves # Also works for 5 and 7
  dplyr::filter(n() !=0)%>%
  dplyr::ungroup(Id)%>%
  dplyr:: mutate(Years = as.numeric(years))%>%
  dplyr::mutate(Age.10yrs = (Age/10))%>%
  dplyr::mutate(Religion.Church.Log = log(Religion.Church+ 1))%>%
  dplyr::mutate(Pol.Orient.S = scale(Pol.Orient, center = TRUE, scale = TRUE),
         Age.10yrs.C = scale(Age.10yrs, center = TRUE, scale = FALSE),
         Male = factor(Gender),
         Employed = factor(Employed),
         Partner = factor(Partner),
         Edu.S = scale(as.numeric(Edu) ,scale=TRUE,center=TRUE),
         Education = as.numeric(Edu),
         Urban = factor(Urban),
         Believe.Spirit = as.factor(Believe.Spirit),
         Believe.God = as.factor(Believe.God),
         EthnicCategories = factor(EthnicCats),
         LIFESAT.S = scale(LIFESAT, center = TRUE, scale = TRUE),
         PWI.S = scale(PWI, center = TRUE, scale = TRUE),
         Deprivation.S = scale(NZdep, scale=TRUE, center=TRUE),
         Relid.C = scale(Relid, scale = TRUE,center=TRUE),
         Religious = as.factor(Religious),
         Religion.Church.Log.C = scale(Religion.Church.Log, center = TRUE, scale = FALSE))
```
```{r}
# nice labels
table1::label(dfben$Age.10yrs.C) <- "Age in Decades (C)"
table1::label(dfben$PWI) <- "Personal Wellbeing"
table1::label(dfben$LIFESAT) <- "Life Satisfaction"

@@

The relationship between spirituality and psychological well-being is a matter of enduring fascination [@Ellsworth2010-yu]. Previous attempts to quantify the relationship between these two domains reveal a positive association: spirituality predicts well-being [@Russinova2002-rq; @Ginsburg1995-jr; @Zaza2005-ac; @Koenig2010-gk; @Smith2003-re; @Smith2003-re; @Koenig2001-ow]. However, critics raise three challenges.

@@

First, previous research has frequently operationalised spirituality using measures of well-being. For example, The Daily Spiritual Experience Scale, which is widely used in spiritual mental health research, includes items such as "I feel thankful for my blessings" and "I feel deep inner peace or harmony" to measure spirituality [@Underwood2002-hg]. As Koenig observes, it is unsurprising that spirituality predict well-being when the same items are used to measure both constructs. Koenig aptly describes this features as a "tautology" in spirituality/mental health research [@Koenig2008-lv]. Indeed, recent systematic-review of spirituality/mental health research found that nearly 45% of previous studies employe tautological scales [@Garssen2016-kb]. The first challenge, then, is to define measures of spirituality that are not well-being measures reclocked in other words.

A second challenge is to disentangle concepts of spirituality from concepts of religion. The term "spirituality" was first used to describe the religious practices of ascetics and monks, however, more recently, the term has taken on a broader family of meanings [@Swinton2001-vr].  However previous research does not converge on a unique set of meanings. Most studies of spirituality and religion have been conducted among (cross-sectional) North American samples among people who are affiliate with traditional religion [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re;  @Visser2010-kq; @Yonker2012-zg]. However, when spirituality has been operationalised as distinct from religion, some researchers have found a negative association within subjective well-being [@King2013-cg]. Thus, in addition to disentangling the concept of spirituality from concepts of subjective well-being, it is also important to disentangel the concept of spirituality from concepts of religion.

A third challenge is to investigate the relationship between spirituality and subjective well-being over the life-span in nationally diverse samples. To assess whether spiritual beliefs lead to growth, stability, or decline in psychological well-being requires comparing requires repeatedly measuring individuals who differ in spirituality, demography, and religion as these people age, and investigating how their perceive their lives to be turning out [@Garssen2016-km].  However, previous investigations of spirituality and wellbeing, including high-quality reviews and meta-analyses, have relied on cross-sectional samples [@Ano2005-hx; @Hackney2003-rs; @Sawatzky2005-rw; @Smith2003-re; @Visser2010-kq; @Yonker2012-zg]. To address the key questions in spirituality psychological health research, then, requires longitudional investigations.

Here, we addresses the three challenges of previous inthe following ways: First, we avoid a tautological elision of "spirituality" and "well-being" by adopting a non-affective measure of spirituality as a believe in a Spirit or Life Force. To  measure subjective well-being we use a well-validated scale for Personal Well-being and a well-validated scale for Life-Satisfaction (see Method). This design avoids measuring spirituality with an indicator of spirituality that is cloacked as psychological well-being, thereby avoiding tautology.  Second, we adopt a four-level categorical indicator for belief-states to clarify differences among people with spiritual, religious, and skeptical beliefs.  We construct this four-level categorical indicator by combining responses to the Spiritual Belief measure with responses to the question: "Do you believe in some form of a spirit or life force?" The levels of our four-level categorical indicator of spiritual and religious beliefs are as follows: (1) Skeptics: those who neither believe in a God nor believe in a Spirit or Life force; (2) Skeptics about God who nevertheless believe in a Spirit or Life Force (henceforth, SpiritExcludeGod); (3) Those who believe in a God but do not believe in a Spirit or Life force (henceforth, GodAndSpirit); (4) Those who who believe in a both God and a Spirit or Life Force, (henceforth GodExcludesSpirit). Narrowing the concept of spirituality to belief state comes at a price. Such an operationalisation, clearly, cannot simultaneously address all interests in previous research. However, this cost to scope is paid for with the benefits of precision. The spiritual indicators measure cognitive states in which distinct varieties of skeptical beliefs, traditional religious beliefs, and non-traditional spirit beliefs can are disentangled. Third, to assess whether spirit beliefs lead to growth, stability, or decline in psychological well-being, we leverage longitudinal responses from a nationally diverse probability sample year over an nine--year period with nine measurement points (2010 to 2018 New Zealand Attitudes and Values Study, NZAVS).

@@

@@

**Personal well-being** Personal well-being is a measure of function all well-being that was assessed using a 4-item version of the Australian Unity well-being Index [@Cummins2017-ur]. Participants rated their satisfaction with (a) "Your standard of living"; (b) "Your health"; (c) "Your future security"; (d) "Your personal relationships." Items were rated using a 10-point response option ranging from 1 = completely dissatisfied to 10 = completely satisfied. The mean Cronbach’s alpha for this scale was $\alpha$ = 0.90. Higher scores on this scale indicate higher personal well-being. Overall scores were means of the 4-items. For detailed information pertaining to means, standard deviations, and missingness, see Table 1.

@@

Our findings support critiques of the reliability of past methods.  Still, further research is needed to determine if other aspects of religion and spirituality, such as practices and rituals, might yield different results.
